---
title: "Project 3. Viterbi algorithm"
date: March 10, 2024
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Problem 6

Consider the HMM represented by the above graph, where $X_n$ are observed variables and $Z_n$ are hidden variables. Each hidden variable can take on $K$ different values and the observed variables can take on $M$ different realisations. The HMM is now parameterized as in the lecture on slide 18.

a) What is the maximum number of free parameters to define the HMM?

Parameters of HMM:

- Initial state probabilities: $I_k=P(Z_1 = k)$ -> vector of length k, probabilities sum up to 1 so k-1 free parameters

- Transition probabilities: $T_{kl}=P(Z_n = l | Z_{n-1} = k)$ from state k to l. Generally, it would be a matrix of size $k \times k$ (from any state k to any other state), with $k \cdot (k-1)$ free parameters

- Emission probabilities: $E_{kx} = P(X_n = x | Z_n = k)$. With k states and m random variables x, thus a matrix of size  $k \times m$ , with $k \cdot (m-1)$ free parameters

The overall sum of parameters is: (k-1) + (k * (k-1)) + (k * (m-1))

b) Assume K = 2 and the transition matrix is

\begin{equation}
T=
\begin{pmatrix}
0.3 & 0.7\\
0.2 & 0.8
\end{pmatrix}
\end{equation}

What is the stationary distribution $\pi$?

The definition of stationary distribution satisfies the following:

$$
\pi = \pi  T
$$

Considering K is equal to 2, that means that the stationary distribution can be represented as following vector with k-1=1 parameters:

\begin{equation}
\pi=
\begin{bmatrix}
p & (1-p)
\end{bmatrix}
\end{equation}


This leads to the following equations:

$$
0.3p + 0.2(1-p) = p
$$
$$
0.7p + 0.8(1-p) = 1-p
$$
Both this equations lead to the solutions of $p=\frac{2}{9}$. Thus the stationary distribution is:


\begin{equation}
\pi=
\begin{bmatrix}
\frac{2}{9} &,     \frac{7}{9}
\end{bmatrix}
\end{equation}

# Problem 7

## Setting up the environment

```{r}
suppressPackageStartupMessages(library(dplyr))
set.seed(42)
```
```{r}
DSSP <- c("B", "C", "E", "G", "H", "I", "S", "T")
AA <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", 
    "L", "M", "N", "P", "Q", "R", "S", "T", "U", 
    "V", "W", "X", "Y"
)
data_folder <- "../data/"
```

### Functions

Viterbi algorithm was provided.

```{r}
#' @param E n times m matrix with the emission log probabilities of the n latent variables and m observed variables
#' @param Tr n times n matrix with the transition log probabilities
#' @param I numeric vector of length n with the initial log probabilities of the n latent variables
#' @param p a data.frame with a Variable named AminoAcids which holds the amino acid sequence as a character string
viterbi <- function(E, Tr, I, p) {
    .as.array <- function(.) stringr::str_split(., "")[[1]]
    unique.ss <- c("B", "C", "E", "G", "H", "I", "S", "T")
    unique.aa <- c("A", "C", "D", "E", "F", "G", "H", "I",
                   "K", "L", "M", "N", "P", "Q", "R", "S",
                   "T", "U", "V", "W", "X", "Y")
    for (k in seq(nrow(p))) {
        sequence <- p$AminoAcids[k]
        aa.vec <- .as.array(sequence) %>% match(unique.aa)
        P      <- matrix(0, nrow(E), length(aa.vec))
        Ptr    <- matrix(0, nrow(E), length(aa.vec))
        
        ## sets the paths
        for (i in seq(length(aa.vec))) {
            if (i == 1) {
                P[, i] <- I + E[, aa.vec[i]]
            } else {
                for (j in seq(nrow(E))) {
                    p.loc    <- P[, i - 1] + Tr[, j] + E[j, aa.vec[i]]
                    P[j, i] <- max(p.loc)
                    Ptr[j, i] <- which.max(p.loc)
                }
            }
        }
        
        ## backtrace: computes the most likely path
        Phi <- vector(mode="integer",   length=length(aa.vec))
        Phi[length(Phi)] <- which.max(P[, ncol(P)])
        ## we start at the back, just as with Needleman-Wunsch or Smith-Waterman
        for (i in seq(from=length(aa.vec), to=2)) {
            Phi[i - 1] <- Ptr[Phi[i], i]
        }
        
        states <- unique.ss[Phi]
        p$PredictedStructure[k] <- paste(states, collapse="")
    }
    return(p)
}

loading_data <- function(data_file, sec_struct) {
    col_names <- c('SeqID', 'AminoAcids')
    if(sec_struct) { col_names <- append(col_names, 'SecondaryStructure')}
    data_df <- read.csv(data_file, header=FALSE, sep="\t")
    names(data_df) <- col_names
    return(data_df)
}

assign_row_names <- function(M) {
    row.names(M) <- DSSP
    return(M)
}

assign_col_names <- function(M, aa=TRUE) {
    if (aa) {
        colnames(M) <- AA
    } else {
        colnames(M) <- DSSP
    }
    
    return(M)
}

init_matrices <- function(data) {
    # Initialisation of I, T, and E matrices using maximum likelihood

    # I - 8 x 1 matrix - initial state probabilities
    I <- matrix(0, nrow=length(DSSP), ncol=1)
    # T - 8 x 8 matrix - transition probabilities
    T <- matrix(0, nrow=length(DSSP), ncol=length(DSSP))
    # E - 8 x 22 matrix - emission probabilities
    E <- matrix(0, nrow=length(DSSP), ncol=length(AA))

    I <- assign_row_names(I)
    T <- assign_row_names(T)
    E <- assign_row_names(E)

    T <- assign_col_names(T, aa=FALSE)
    E <- assign_col_names(E)

    I <- initial_states(data, I)
    T <- transition_states(data, T)
    E <- emission_states(data, E)

    I <- log(I)
    T <- log(T)
    E <- log(E)

    return(list(I=I, T=T, E=E))
}

initial_states <- function(data, M) {
    # Retrieving DSSP profiles
    sec_struct <- data$SecondaryStructure
    
    # Computing frequencies of each letter in the first position
    frequency <- table(substr(sec_struct, 1, 1))

    # Computing relative frequencies
    M[row.names(M) %in% names(frequency), 1] <- as.numeric(frequency)/sum(frequency)

    # Return initial states matrix
    return(M)
}

transition_states <- function(data, M) {
    # Retrieving DSSP profiles
    sec_struct <- data$SecondaryStructure
    
    pairs <- lapply(sec_struct, 
        function(x) substring(x, first=1:(nchar(x)-1), last=2:nchar(x))
    )
    frequencies <- table(unlist(pairs))

    # Assigning frequencies
    for (i in seq(nrow(M))) {
        for (j in seq(ncol(M))) {
            row_name <- row.names(M)[i]
            col_name <- colnames(M)[j]
            freq_name <- paste0(row_name, col_name)
            if (is.na(frequencies[freq_name])) {
                M[i, j] <- 1
            } else {
                M[i, j] <- as.numeric(frequencies[freq_name])+1
            }
        }
    }
    # Relative frequencies
    M <- M/rowSums(M)

    # Return transition states matrix
    return(M)
}

emission_states <- function(data, M) {
    # Retrieving profiles
    aa <- data$AminoAcids
    dssp <- data$SecondaryStructure

    aa_list <- lapply(aa,
        function(x) substring(x, first=1:(nchar(x)), last=1:nchar(x))
    )
    dssp_list <- lapply(dssp, 
        function(x) substring(x, first=1:(nchar(x)), last=1:nchar(x))
    )
    
    frequencies <- table(paste(unlist(dssp_list), unlist(aa_list), sep=""))

    # Assigning frequencies
    for (i in seq(nrow(M))) {
        for (j in seq(ncol(M))) {
            row_name <- row.names(M)[i]
            col_name <- colnames(M)[j]
            freq_name <- paste0(row_name, col_name)
            if (is.na(frequencies[freq_name])) {
                M[i, j] <- 1
            } else {
                M[i, j] <- as.numeric(frequencies[freq_name])+1
            }
        }
    }
    # Relative frequencies
    M <- M/rowSums(M)

    # Return initial states matrix
    return(M)
}

eigen_stationary_distribution <- function(T) {
    # T transposed to get left eigenvectors of T
    e <- eigen(t(exp(T)))
    e_vec <- e$vectors
    e_val <- e$values
    e_val_one_index <- which(abs(e_val-1) < 1e-12)
    stat_dist <- matrix(e_vec[, e_val_one_index]/
        sum(e_vec[, e_val_one_index]), nrow=1, ncol=length(DSSP)
    )
    colnames(stat_dist) <- DSSP
    print(paste("Eigenvalue:", e_val[e_val_one_index]))
    print("Stationary distribution of transmission probabilities (eigenvalue approach): ")
    return(stat_dist)
}

brute_force_stationary_distribution <- function(T) {
    T <- exp(T)
    while (sum(abs(T[1,]-T[2,])) > 1e-12) {
        T <- T %*% T
    }
    print("Stationary distribution of transmission probabilities (brute-force approach): ")
    print(T[1,])
}

run_viterbi_predictions <- function(data, params) {
    pred_df <- data.frame(AminoAcids=data$AminoAcids, PredictedStructure=NA)
    pred_df <- apply(pred_df, 1, function(row) {
        viterbi(params$E, params$T, params$I, data.frame(AminoAcids=row["AminoAcids"]))
    })
    pred_df <- data.frame(t(sapply(pred_df, function(x) x[1:max(lengths(pred_df))])))
    data$PredictedStructure <- pred_df$PredictedStructure
    data <- apply(data, 2, as.character)
    return(data)
}

run_random_predictions <- function(data, params) {
    rand_struct <- apply(data, 1, function(row) {
        paste0(sample(DSSP, nchar(row["AminoAcids"]), replace=TRUE), collapse="")
    })
    data <- cbind(data, RandomStructure=rand_struct)
    return(data)
}

save_to_tsv <- function(data, file_path) {
    write.table(data, file=file_path, sep="\t", row.names=FALSE, col.names=FALSE)
}

get_accuracies <- function(pred_df, column_name) {
    accuracies <- apply(pred_df, 1, function(row) {
        sum(
            strsplit(row["SecondaryStructure"], "")[[1]] == strsplit(row[column_name], "")[[1]])/
            nchar(row["SecondaryStructure"])
    })
    print(summary(accuracies))
    return(accuracies)
}

```

## (a) Data loading 

Read `proteins_train.tsv`, `proteins_test.tsv` and `proteins_new.tsv` into the memory and store each in a data.frame.

```{r}
# Loading data
prot_train_df <- loading_data(paste0(data_folder, "proteins_train.tsv"), sec_struct=TRUE)
prot_test_df <- loading_data(paste0(data_folder, "proteins_test.tsv"), sec_struct=TRUE)
prot_new_df <- loading_data(paste0(data_folder, "proteins_new.tsv"), sec_struct=FALSE)
```

## (b) Estimating parameters

Estimate the vector of initial state probabilities I, the matrix of transition probabilities T and the matrix for emission probabilities E by using maximum likelihood, i.e., by counting how often each transition and each emission happened and how often each secondary structure was found at the beginning of a protein, respectively.

$$
I_i = P(Z_1 = i) = \frac{\#latent \; sequences \; start \; with \;  i}{\# sequences}
$$

$$
E_{ij} = P(X_n=j|Z_n = i) = \frac{\#latent \; state \; i \; emits \; observable \;j}{\# latent \; state \; i \; emits \; anything}
$$

$$
T_{ij} = P(Z_n=j|Z_{n-1} = i) = \frac{\#latent \; state \; i \; transits \; to \; state \;j}{\# latent \; state \; i \; transits \; to \; any \; other \; state}
$$
```{r}
# Initialisation of parameters
params <- init_matrices(prot_train_df)
```

# (c) Stationary distribution

## Eigenvalues 

We need left eigenvector for the stationary distribution:

$$
\pi = \pi  T
$$

```{r}
eigen_stationary_distribution(params$T)
```

## Brute force 

```{r}
brute_force_stationary_distribution(params$T)
```

# (d) Predictions

```{r}
# Making predictions
prot_test_df <- run_viterbi_predictions(prot_test_df, params)
prot_new_df <- run_viterbi_predictions(prot_new_df, params)

# Saving predictions
save_to_tsv(prot_new_df, paste0(data_folder, "proteins_new_pred.tsv"))

# Computing prediction accuracies
print("Accuracy statistics of Viterbi predictions:")
viterbi_acc <- get_accuracies(prot_test_df, column_name="PredictedStructure")

# Making random "predictions"
prot_test_df <- run_random_predictions(prot_test_df, params)
print("Accuracy statistics of random 'predictions':")
rand_acc <- get_accuracies(prot_test_df, column_name="RandomStructure")
```

# (e) Bootstrapping

```{r}
# Number of bootstrap runs
n_bt <- 1000
# Define storage dataframes
I_bt <- matrix(0, length(DSSP), n_bt)
E_bt <- array(0, dim=c(length(DSSP), length(AA), n_bt))
T_bt <- array(0, dim=c(length(DSSP), length(DSSP), n_bt))

# Bootstrapping
for (i in 1:n_bt){
    # Resample the dataframe
    df_bt <- prot_train_df[sample(nrow(prot_train_df), replace=T),]
    
    est <- init_matrices(df_bt)

    # Estimate current run
    I_est <- exp(est$I)
    E_est <- exp(est$E)
    T_est <- exp(est$T)
    
    # Concatenate
    I_bt[, i] <- I_est
    E_bt[, , i] <- E_est
    T_bt[, , i] <- T_est
    
    if (sum(is.na(T_est)) > 0){
        break
    }
}
```

```{r}
I_ci <- apply(I_bt, 1, function(x) {
  down <- quantile(x, 0.025)
  up <- quantile(x, 0.975)
  return(c(down, up))
})

E_ci <- apply(E_bt, c(1, 2), function(x) {
  down <- quantile(x, 0.025)
  up <- quantile(x, 0.975)
  return(c(down, up))
})

T_ci <- apply(T_bt, c(1, 2), function(x) {
  down <- quantile(x, 0.025)
  up <- quantile(x, 0.975)
  return(c(down, up))
})

I_ci
head(E_ci)
head(T_ci)
```

# (g) Comparison

```{r}
# Plot box plot of accuracies
boxplot(rand_acc, viterbi_acc, ylim=c(0, 1), ylab="Accuracy", xlab="Algorithm",
    names=c("Random", "Viterbi")
)
```

## Render this .rmd into a pdf

```{r, eval=FALSE}
library(rmarkdown)
render("viterbi.Rmd", pdf_document(TRUE), "viterbi.pdf")
```
