---
title: "Project 7"
output:
  pdf_document: 
    extra_dependencies: ['amsmath']
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

library(foreach)
library(doParallel)
```

# Problem 19
## Message passing on a chain

We will now use R to perform the message passing shown in Figure 1. Suppose that we have the following (conditional) probability distributions:

```{r}
# Conditional probabilities
x1 <- 1/3
x2_1 <- c(4/5, 2/3)
x3_2 <- c(5/7, 1/3)
x4_3 <- c(3/5,  2/5)
x5_4 <- c(1/2, 7/9)
```

Note that these equations fully determine each (conditional) probability distribution, since $X_i \in \{0, 1\}$ for $i \in \{1, . . . , 5\}$.

## (a) Store clique potentials in an R object

```{r}
# Potential matrix
pot <- array(dim = c(2, 2, 4), dimnames = list(c("0", "1"), c("0", "1"),
c("Psi12", "Psi23", "Psi34", "Psi45")))

# Filling it up
pot[, , "Psi12"] <- cbind(1-x2_1, x2_1) * c(1-x1,x1)
pot[, , "Psi23"] <- cbind(1-x3_2, x3_2) 
pot[, , "Psi34"] <- cbind(1-x4_3, x4_3) 
pot[, , "Psi45"] <- cbind(1-x5_4, x5_4) 

pot
```

## (b) Computing forward messages

```{r}
# Forward message
mu_a <- array(dim = c(5, 2), dimnames = list(c("X1", "X2", "X3", "X4", "X5"), c("0", "1")))

# Initialization
mu_a[1, ] <- 1

# Computation
for (i in 2:5){
  mu_a[i, ] <- mu_a[i-1, ] %*% pot[, , i-1]
}

mu_a
```

## (c) Computing backward message

```{r}
# Backward message
mu_b <- array(dim = c(5, 2), dimnames = list(c("X1", "X2", "X3", "X4", "X5"), c("0", "1")))

# Initialization
mu_b[5, ] <- 1

# Computation
for (i in 4:1){
  mu_b[i, ] <- mu_b[i+1, ] %*% pot[, , i]
}

mu_b
```

## (d)) Compute the marginal probability distribution for each no

```{r}
# Marginal prob.
marg <- array(dim = c(5, 2), dimnames = list(c("X1", "X2", "X3", "X4", "X5"), c("0", "1")))

# Calculate
for (i in 1:5){
  marg[i, ] <- mu_a[i, ] * t(mu_b)[, i]
}

marg
```

### Normalizing constant Z

```{r}
Z <- rowSums(marg)
Z
```


### Normalizing
```{r}
marg <- marg/Z
print(rowSums(marg))
print(marg)
```

```{r, eval=FALSE}
library(rmarkdown)
render("ex7.Rmd", pdf_document(TRUE), "Indilewitsch_Toidze_Houhamdi_Pudziuvelyte_Project7.pdf")
```