---
title: "Basic R tutorial"
date: February 22, 2024
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Problem 3: Learning Bayesian networks from protein data

## Preparation to run the code

Setting seed for reproducibility and loading packages.

```{r}
set.seed(42)
library("GGally")
library("BiDAG")
```

## Loading data

```{r}
data <- read.table("2005_sachs_2_cd3cd28icam2_log_std.csv", sep=",", header=TRUE)
```

## Variables and observations

```{r}
num_variables <- ncol(data)
num_observations <- nrow(data)
```

## Visualisation of transformed data

TODO: must resize the plot

```{r}
pair_plot <- ggpairs(data, progress=FALSE)
pair_plot
```

TODO: execute the following routine with different parameters (10 times for each change of parameter)

TODO: include possibility of parallelization

## Defining functions

```{r}
splitting_data <- function(data) {
    # Shuffling data
    indices <- 1:nrow(data)
    indices <- sample(length(indices))

    # Splitting data
    train_size <- ceiling(nrow(data)*0.8)
    train_indices <- indices[1:train_size]
    test_indices <- indices[(train_size+1):(length(indices))]

    # Checking if there is no overlap
    if(length(unique(c(train_indices, test_indices))) != length(c(train_indices, test_indices))) {
        print("Overlap!")
    }

    train_data <- data[row.names(data) %in% train_indices, ]
    test_data <- data[row.names(data) %in% test_indices, ]

    split_data <- list("train_data"=train_data, "test_data"=test_data)

    return(split_data)
}

training_BN <- function(data, bgepar) {
    init_score_par <- scoreparameters("bge", data$train_data, bgepar)
    learnt_BN <- iterativeMCMC(init_score_par, verbose=FALSE)

    return(learnt_BN)
}

testing_BN <- function(data, BN, bgepar) {
    test_score_par <- scoreparameters("bge", data$test_data, bgepar)
    test_score <- scoreagainstDAG(test_score_par, BN$DAG)
    return(mean(test_score))
}
```

## Default parameters

```{r}
bgepar <- list(am=1, aw=NULL, edgepf=1)
split_data <- splitting_data(data)
learnt_BN <- training_BN(split_data, bgepar)
mean_test_score <- testing_BN(split_data, learnt_BN, bgepar)
print(paste0("Average BGe score on testing data: ", mean_test_score))
```

## Different parameters

```{r}
ams <- c(10^(-5), 10^(-3), 10^(-1), 10, 10^2)

for(am in ams) {
    print(paste0("Running BN training with a_m: ", am))
    bgepar <- list(am=am, aw=NULL, edgepf=1)
    split_data <- splitting_data(data)
    learnt_BN <- training_BN(split_data, bgepar)
    mean_test_score <- testing_BN(split_data, learnt_BN, bgepar)
    print(paste0("Average BGe score on testing data: ", mean_test_score))
}
```

## Render this .rmd into a pdf

```{r, eval=FALSE}
library(rmarkdown)
render("1/1.Rmd", pdf_document(TRUE), "1.pdf") # TRUE adds table of content
```
